---
title: "fameR_Rmarkdown"
author: "FAE"
date: "2026-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Notes for GitHub users:**

If you cloned this repository, you can **knit this R Markdown file exactly as it is** to run the complete pipeline using the **example dataset included in the repo**.

The example output will be generated inside the `example_output/` folder.

Make sure the `fameR.Rmd` file is opened inside the **`fameR_Rmarkdown.Rproj`** so that relative paths resolve correctly.

We aim to convert this pipeline into an R package in the near future.


```{r libraries}
library(tidyverse)
library(readxl)
library(stringr)
library(ggpmisc)
library(openxlsx)
library(tools)
library(writexl)
library(ggpubr)
library(rstatix)
library(scales)
```


# FUNCTIONS (Not intended for user modification - unless necessary)*

Clean - Function for initial cleaning unformatted GC-FID data:
```{r message=FALSE, warning=FALSE, include=FALSE}
Clean <- function(RAWDATA) {
  #Clean line breaks in column names
  names(RAWDATA) <- gsub("[\r\n]+","", names(RAWDATA))
  
  #Clean line breaks from all character columns
  RAWDATA[] <- lapply(RAWDATA, function(col) {
    if (is.character(col)) {
      gsub("[\r\n]+","", col)
    } else {
      col
    }
  })
  
  names(RAWDATA)[names(RAWDATA) == "Sample Description"] <- "Experiment_ID"
  
  return(RAWDATA)
}
```

ProcessGCFiles - Function to import GC-FID data in one folder (uses Clean function):
```{r message=FALSE, warning=FALSE, include=FALSE}
ProcessGCFiles <- function(folder_path, verbose = TRUE) {
  # List all Excel files in the folder
  GCFID_files <- list.files(folder_path, pattern = "\\.XLSX$", full.names = TRUE, ignore.case = TRUE)
  
  # Internal function to process each file
  process_fame_file <- function(path) {
    sequence_id <- tools::file_path_sans_ext(basename(path))
    name <- paste0(sequence_id, "_i")
    
    if (verbose) message(paste("📥 Imported GC-FID run:", sequence_id))
    
    tryCatch({
      RAWDATA <- readxl::read_excel(path, sheet = "Page 2", n_max = 2000)
      cleaned_data <- Clean(RAWDATA) %>%
        dplyr::rename(Compound = Name) %>%  # Rename 'Name' to 'Compound'
        dplyr::mutate(Sequence_ID = sequence_id)
      setNames(list(cleaned_data), name)
    }, error = function(e) {
      message(paste("❌ Failed to import:", sequence_id, "-", e$message))
      NULL
    })
  }
  
  # Apply to all files and flatten into a named list
  purrr::map(GCFID_files, process_fame_file) %>% purrr::compact() %>% purrr::flatten()
}
```

GenerateLevelMap - Function to map the names of standard samples for each Sequence_ID
```{r}
GenerateLevelMaps <- function(GCFID_Results, cal_path, verbose = TRUE) {
  # Get all unique Sequence_IDs from the list of tibbles
  sequence_ids <- unique(unlist(lapply(GCFID_Results, function(df) unique(df$Sequence_ID))))
  
  # Initialize storage
  level_maps <- list()
  level_maps_df <- list()
  
  # Extract level maps for each Sequence_ID
  for (seq_id in sequence_ids) {
    raw_levels <- suppressMessages(
      readxl::read_excel(
        path = cal_path,
        sheet = seq_id,
        range = "D13:M13",
        col_names = FALSE
    ))
    
    level_ids <- unlist(raw_levels)
    level_ids <- level_ids[!is.na(level_ids) & level_ids != ""]
    
    level_map <- setNames(
      paste0("Level", seq_along(level_ids), " [mg/L]"),
      level_ids
    )
    
    level_maps[[seq_id]] <- level_map
    
    # Build tidy summary for this sequence
    level_maps_df[[seq_id]] <- tibble::tibble(
      Sequence_ID = seq_id,
      Standard_ID = level_ids,
      Level_Label = level_map
    )
    
    # Verbose output
    if (verbose) {
      cat(paste0("\n📘 Sequence_ID: ", seq_id, "\n"))
      cat("🧪 Standards found: ", paste(level_ids, collapse = ", "), "\n")
    }
  }
  
  # Combine all summaries into one tidy data frame
  level_maps_df <- dplyr::bind_rows(level_maps_df)

  # Return both the list and the summary table
  return(list(
    maps = level_maps,
    summary = level_maps_df
  ))
}
```

ImportCalibrationMatrix - Function to import all the external calibrants information based on unique Sequence_IDs taken from the calibrants Level maps
```{r message=FALSE, warning=FALSE, include=FALSE}

ImportCalibrationMatrix <- function(cal_path, GCFID_LevelMaps, verbose = TRUE) {
  sequence_ids <- names(GCFID_LevelMaps)
  all_cal_data <- list()
  
  for (seq_id in sequence_ids) {
    if (verbose) cat(paste0("\n📘 Processing calibration for Sequence_ID: ", seq_id, "\n"))
    
    # Step 1: Read header row (row 13) to get actual standard IDs
    header_row <- suppressMessages(suppressWarnings(
      readxl::read_excel(
        path = cal_path,
        sheet = seq_id,
        range = "D13:M13",
        col_names = FALSE
      )
    ))
    
    std_ids <- unlist(header_row)
    std_ids <- std_ids[!is.na(std_ids) & std_ids != ""]
    
    # Step 2: Read full calibration matrix (row 14 onward)
    cal_data <- suppressMessages(suppressWarnings(
      readxl::read_excel(
        path = cal_path,
        sheet = seq_id,
        range = "A14:M100",
        col_names = FALSE
      )
    ))
    
    # Extract compound names from column A
    compounds <- cal_data[[1]]
    
    # Extract only the columns that match named standards
    all_conc_cols <- cal_data[, 4:(3 + ncol(header_row))]
    
    # Identify which columns correspond to named standards
    named_col_indices <- which(!is.na(header_row) & header_row != "")
    conc_matrix <- all_conc_cols[, named_col_indices]
    
    # Assign correct column names
    colnames(conc_matrix) <- std_ids
    
    # Get level labels from level map
    level_labels <- GCFID_LevelMaps[[seq_id]]
    
    # Build tidy table
    cal_tidy <- tidyr::pivot_longer(
      dplyr::bind_cols(Compound = compounds, conc_matrix),
      cols = -Compound,
      names_to = "Sample Name",
      values_to = "Concentration"
    ) %>%
      dplyr::mutate(
        Sequence_ID = seq_id,
        Level_Label = level_labels[`Sample Name`]
      ) %>%
      dplyr::filter(!is.na(Concentration) & !is.na(Compound))
    
    all_cal_data[[seq_id]] <- cal_tidy
  }
  
  # Combine all into one data frame
  dplyr::bind_rows(all_cal_data)
}

```

AttachGCResultsToCalibration
```{r}

AttachGCResultsToCalibration <- function(GC_calibration, GC_results, IS_name = "C13:0 ISMethyltridecanoate", verbose = TRUE) {
  updated_cal <- GC_calibration %>%
    dplyr::group_split(Sequence_ID) %>%
    purrr::map_dfr(function(cal_group) {
      seq_id <- unique(cal_group$Sequence_ID)
      
      # Match GC_results element by partial Sequence_ID
      matched_name <- names(GC_results)[stringr::str_detect(names(GC_results), fixed(seq_id))]
      gc_data <- if (length(matched_name) == 1) GC_results[[matched_name]] else NULL
      
      if (is.null(gc_data)) {
        if (verbose) cat(paste0("⚠️ No GC-FID data found for Sequence_ID: ", seq_id, "\n"))
        return(cal_group)
      }
      
      # Rename for clarity
      gc_data <- gc_data %>% dplyr::rename(Compound = Compound)  # Just in case
      
      # Join calibration with GC-FID areas by Sample Name and Compound
      cal_with_area <- dplyr::left_join(
        cal_group,
        gc_data %>% dplyr::select(`Sample Name`, Compound, Area),
        by = c("Sample Name", "Compound")
      )
      
      # Extract IS areas per Sample Name
      IS_areas <- gc_data %>%
        dplyr::filter(Compound == IS_name) %>%
        dplyr::select(`Sample Name`, IS_Area = Area)
      
      # Join IS areas to calibration data
      cal_with_area <- dplyr::left_join(
        cal_with_area,
        IS_areas,
        by = "Sample Name"
      )
      if (verbose) {
        if (nrow(IS_areas) > 0) {
          cat(paste0("\n🧪 Found IS areas for ", nrow(IS_areas), " samples and standards in Sequence_ID: ", seq_id, "\n"))
        } else {
          cat(paste0("⚠️ Internal Standard '", IS_name, "' not found in Sequence_ID: ", seq_id, "\n"))
        }
      }
      
      return(cal_with_area)
    })
  
  return(updated_cal)
}
```


# GENERAL INSTRUCTIONS FOR NEW DATA PROCESSING

1) Acquire and fill-out all the data file templates (GC-FID report template, Metadata template, FAME standard template, and Quantification rules template - The easiest way would be to modify the example datasets). Ask questions to Fuad (fuadale@lanl.gov) or relevant point of contact.


2) If not running with example data from the GitHub repo, make sure you have four items in your computer ready to import:

    a) GC-FID integration reports exported from ChemStation (or equivalent GC-FID report template). These files should be all in a single folder and named EXACTLY after the "Sequence_ID" column in the Sample metadata information file. Note that if the correct template was used, these files have to also contain a column with "Experiment_ID" which should exactly match the "Experiment_ID" column in the Sample metadata information file.
    
    b) Sample metadata information file with the weight of the samples and relevant experimental information (use the FAME meta.xlsx template). Please note that this version of fameR does not handle analytical replicates (for example; samples that were injected multiple times in the GC instrument or that received the same Injection_ID).
    
    c) Updated full profile of external standards (use FAME standard levels.xlsx template) with an additional sheet specifying the relationships between the compound names as exported in the GC reports (often short or not very good looking names) and the desired labels as intended to be shown in your output dataset and plots. Note that this pipeline follows common analytical chemistry practices for quantification using an internal standard and won't work well if the data was collected in conditions that don't follow these common practices. If your GC runs did not include an internal standard, this pipeline needs to be modified to conduct only external calibration.
    
    d) OPTIONAL BUT ENCOURAGED: If you conducted identification on GC-MS and/or have quantification rules for uncalibrated compounds, you will need a Quantification rules file that states the surrogate external standard compounds which response factor will be used to calculate the amount of each uncalibrated compound. Note that the best approach for this is to have the uncalibrated peaks identified in ChemStation (or equivalent chromatography data acquisition software), in such a way that they already appear as identified uncalibrated peaks in the GC-FID integration reports. In the example data the uncalibrated peaks are marked with (MS) at the end of the compound names. The surrogate peaks selection for quantification of uncalibrated compounds in the example set already follow standard rules based on structural similarity (https://docs.nrel.gov/docs/fy16osti/60958.pdf). The same rules are recommended for other cases.


3) All the user inputs that need to be updated each time that this pipeline is newly set-up are indicated with empty brackets [  ]. Once you have made the input and are sure everything is correct place an X in between the brackets [ X ] to keep control of what parts of the user inputs are already fulfilled.


4) Run once all the chunks above to make sure that all functions are defined and all libraries are loaded. Make sure to install any missing packages.


5) Provide the location of the folder containing the GC-FID integration report on **user_ReportGC** (don't forget to run the chunk). Leave as is if you cloned the GitHub repo and you are working with the provided example data:

USER INPUT [  ] 
```{r}
#USER INPUT
user_ReportsGC <- "your path to GC-FID reports folder..."



# Default path for users who cloned the GitHub repo and are working with example data
default_path <- file.path(getwd(), "example_data", "GC reports")

# Uses user path ONLY if the folder exists; otherwise, falls back to the example data
GCFID_data_path <- if (dir.exists(user_ReportsGC)) {
  user_ReportsGC
} else {
  message("Using example GC-FID data bundled with the fameR GitHub repository.")
  default_path
}

GCFID_data_path
```


6) Provide the location of the file containing all the calibrants information **user_calibration_path**, this file has one tab per Sequence_ID. Update the quantification rules for uncalibrated FAME **user_quant_path**. Don't forget to add the file extensions. It is highly recommended to use the example files as templates. Leave as is if you cloned the GitHub repo and you are working with the provided example data:

USER INPUT [  ]
```{r}
#USER INPUT
user_calibration_path <- "your path to calibration information spreadsheet... use of example template highly recommended... .xslx"
user_calibration_path_labelssheet <- "name of your labels sheet..."

user_quant_path       <- "your path to quantification rules spreadsheet... use of example template highly recommended... .xslx"
user_quant_path_sheet <- "name of your rules sheet..."



# Default path for users who cloned the GitHub repo and are working with example data
default_calibration_path <- file.path(getwd(), "example_data", "FAME standard levels.xlsx")
default_calibration_path_labelssheet <- "Labels"

default_quant_path       <- file.path(getwd(), "example_data", "Uncalibrated.xlsx")
default_quant_path_sheet <- "Rules"

# Uses user paths ONLY if the files exist; otherwise, falls back to the example data
GCFID_calibration_path <- if (file.exists(user_calibration_path)) {
  user_calibration_path
} else {
  message("Using example calibration data bundled with the fameR GitHub repository.")
  default_calibration_path
}

Compounds_labels_sheet <- if (file.exists(user_calibration_path)) {
  user_calibration_path_labelssheet
} else {
  default_calibration_path_labelssheet
}

GCFID_quant_path <- if (file.exists(user_quant_path)) {
  user_quant_path
} else {
  message("Using example quantification rules bundled with the fameR GitHub repository.")
  default_quant_path
}

GCFID_quant_sheet <- if (file.exists(user_quant_path)) {
  user_quant_path_sheet
} else {
  default_quant_path_sheet
}
```


7) Update the location of the file containing all the experiments metadata (use of example template highly recommended). Leave as is if you cloned the GitHub repo and you are working with the provided example data:

USER INPUTS [  ]
```{r}
#USER INPUT
user_meta_path <- "your path to metadata samples information spreadsheet... use of example template highly recommended... .xslx"
user_meta_path_sheet <- "name of your metadata sheet..."



# Default path for users who cloned the GitHub repo and are working with example data
default_meta_path <- file.path(getwd(), "example_data", "FAME meta.xlsx")
default_meta_path_sheet <- "FAME Meta"

# Uses user paths ONLY if the files exist; otherwise, falls back to the example data
GCFID_meta_path <- if (file.exists(user_meta_path)) {
  user_meta_path
} else {
  message("Using example metadata bundled with the fameR GitHub repository.")
  default_meta_path
}

GCFID_meta_path_sheet <- if (file.exists(user_meta_path)) {
  user_meta_path_ssheet
} else {
  default_meta_path_sheet
}
```


8) Update the location of the folder where you want your exported results to go. Leave as is if you cloned the GitHub repo and you are working with the provided example data:

USER INPUTS [X]
```{r}
#USER INPUT: Define the export directory
user_export_dir <- "your export directory folder..."



# Default path for users who cloned the GitHub repo and are working with example data
default_export_dir <- file.path(getwd(), "example_output")

invalid_user_dir <- (
  is.null(user_export_dir) ||
  user_export_dir == "" ||
  grepl("your export directory folder", user_export_dir, ignore.case = TRUE) ||
  !dir.exists(dirname(user_export_dir))
)

# If the user directory is usable, uses it; otherwise falls back to example_output
if (invalid_user_dir) {
  message("Using example output directory included in the GitHub repository.")
  Export_dir <- default_export_dir
} else {
  Export_dir <- user_export_dir
}

# ---- ENSURE DIRECTORY EXISTS ----
if (!dir.exists(Export_dir)) {
  dir.create(Export_dir, recursive = TRUE)
  message("📁 Created export folder: ", Export_dir)
} else {
  message("📁 Export folder already exists: ", Export_dir)
}

Export_dir
```


9) Run the chunks below one by one, paying attention to any errors or alerts, if any pops ask Fuad.


# IMPORTING RAW DATA AND INITIAL CLEANING

## Importing GC-FID results
```{r}

if (length(setdiff(list.files(GCFID_data_path, pattern = "(?i)\\.xlsx$", full.names = TRUE),
                   list.files(GCFID_data_path, pattern = "~\\$", full.names = TRUE))) == 0) {
  message("⚠️ No Excel files found in the folder.")
} else {
  GCFID_Results <- ProcessGCFiles(GCFID_data_path)
}


#View(GCFID_Results)
```

## Importing Metadata and calibrant identifiers
```{r}
#Importing metadata
GCFID_meta <- read_excel(GCFID_meta_path, 
    sheet = GCFID_meta_path_sheet, range = "A2:P20000", 
    na = "NA")

#Mapping the standard levels in each run
GCFID_Levels <- GenerateLevelMaps(GCFID_Results, cal_path = GCFID_calibration_path)
GCFID_Levels_map <- GCFID_Levels$maps
GCFID_Levels_df <- GCFID_Levels$summary
```

## Importing External calibrations and matching with internal standard
```{r}
GCFID_calibration <- ImportCalibrationMatrix(cal_path = GCFID_calibration_path, GCFID_LevelMaps = GCFID_Levels_map)

GCFID_calibration <- AttachGCResultsToCalibration(GC_calibration = GCFID_calibration, GC_results = GCFID_Results)

```

## Importing quantification rules for uncalibrated compounds & label rules
```{r}
# Import and clean quantification rules
GCFID_quant <- read_excel(
  GCFID_quant_path, 
  sheet = GCFID_quant_sheet, 
  range = "A1:C100", 
  na = "NA"
) %>%
  na.omit()

# Check if data was imported
if (!is.null(GCFID_quant) && nrow(GCFID_quant) > 0) {
  message("✅ Quantification rules imported successfully: ", nrow(GCFID_quant), " uncalibrated compounds")
  
  # Identify included compounds
  included <- GCFID_quant %>%
    filter(tolower(Exclude) != "yes") %>%
    pull(Compound)
  
  if (length(included) > 0) {
    message("✅ ", length(included), " compounds will be included for quantification:")
    message("   • ", paste(included, collapse = "\n   • "))
  } else {
    message("⚠️ No compounds available for quantification. All are marked for exclusion.")
  }
  
} else {
  message("⚠️ Quantification rules not imported or empty. If you indend to quantify uncalibrated compounds, please check the file path and sheet name")
}

# Import label rules
GCFID_labels <- read_excel(
  GCFID_calibration_path,
  sheet = Compounds_labels_sheet,
  range = "A1:B100",
  na = "NA"
) %>%
  na.omit()
```



# CALIBRATION CURVES

## Calculating Area ratios in calibration table for each Sequence_ID
```{r}
GCFID_calibration <- GCFID_calibration %>%
  mutate(
    Area_ratio = Area / IS_Area
  )
```


## (OPTIONAL) Adding (0,0) points
```{r}
zero_points <- GCFID_calibration %>%
  ungroup() %>%
  distinct(Sequence_ID, Compound) %>%
  mutate(
    Area = 0,
    IS_Area = NA,
    Concentration = 0,
    Level_Label = "Level0 [mg/L]",
    Area_ratio = 0
  )


GCFID_calibration <- GCFID_calibration %>%
  group_by(Sequence_ID, Compound) %>%
  bind_rows(zero_points) %>%
  ungroup()
  
```


## Creating and exporting calibration curves
Once the calibration curves are exported, examine them and update the calibration file if you decide to remove any of the standard levels (Aim for near 0.999 R2). Once updated re Run All Chunks Above.
USER INPUT [x]
```{r warning=FALSE}
# Loop through each Sequence_ID and generate a plot
unique_sequences <- unique(GCFID_calibration$Sequence_ID)

for (seq_id in unique_sequences) {
  # Filter data for this sequence
  seq_data <- GCFID_calibration %>% filter(Sequence_ID == seq_id)
  
  # Create the plot
  seq_plot <- ggplot(seq_data, aes(x = Concentration, y = Area_ratio)) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    
    # Equation in top-left
    stat_poly_eq(
      aes(label = ..eq.label..),
      formula = y ~ x,
      parse = TRUE,
      size = 2,
      label.x.npc = "left",
      label.y.npc = "top"
    ) +
    
    # R² in bottom-right
    stat_poly_eq(
      aes(label = ..rr.label..),
      formula = y ~ x,
      parse = TRUE,
      size = 3,
      label.x.npc = "right",
      label.y.npc = "bottom",
      vjust = 4.5,
      hjust = -1,
      rr.digits = 3
    ) +
    
    facet_wrap(~ Compound, scales = "free") +
    theme_minimal(base_size = 12) +
    theme(
      strip.text = element_text(size = 6.5, hjust = 1)  # Smaller, right-aligned facet labels
    ) +
    labs(
      title = paste("GC-FID Calibration Curves -", seq_id),
      x = "Concentration",
      y = "Area Ratio"
    )
  
  # Export the plot
  ggsave(
    filename = file.path(Export_dir, paste0("GC_calibration_", seq_id, ".jpeg")),
    plot = seq_plot,
    width = 11,
    height = 7,
    dpi = 300
  )
  
  print(paste0("📤 Exported calibration curves plot for Sequence_ID: ", seq_id))
}

```

## Extracting calibration parameters (internal calibration)
```{r}
#Fit models per Sequence_ID and Compound
model_df <- GCFID_calibration %>%
  group_by(Sequence_ID, Compound) %>%
  summarise(
    model = list(lm(Area_ratio ~ Concentration, data = pick(Area_ratio, Concentration))),
    .groups = "drop"
  )

#Extract slope, intercept, and R²
calibration_params <- model_df %>%
  mutate(
    Slope = map_dbl(model, ~ coef(.x)[["Concentration"]]),
    Intercept = map_dbl(model, ~ coef(.x)[["(Intercept)"]]),
    R_squared = map_dbl(model, ~ summary(.x)$r.squared)
  ) %>%
  select(Sequence_ID, Compound, Slope, Intercept, R_squared) %>%
  filter(!is.na(Slope))
```


# QUANTIFICATION OF ANALYTES

## Calculating Area ratios in samples
```{r}
GCFID_Results_filtered <- map(GCFID_Results, function(df) {
  df_filtered <- df %>%
    semi_join(GCFID_meta, by = c(
      "Sequence_ID" = "Sequence_ID",
      "Experiment_ID" = "Experiment_ID",
      "Sample Name" = "Injection_ID"
    ))

  IS_values <- df_filtered %>%
    filter(Compound == "C13:0 ISMethyltridecanoate") %>%
    select(`Sample Name`, IS_Area = Area)

  df_filtered %>%
    left_join(IS_values, by = "Sample Name") %>%
    mutate(Area_ratio = Area / IS_Area) %>%
    filter(Compound != "C13:0 ISMethyltridecanoate")
})
```


## Calculating concentrations for calibrated compounds
```{r}
GCFID_Results_calibrated <- map(GCFID_Results_filtered, function(df) {
  df %>%
    left_join(calibration_params, by = c("Compound", "Sequence_ID")) %>%
    mutate(Concentration_mgperL = (Area_ratio - Intercept) / Slope)
})
```


## Calculating concentrations for uncalibrated compounds
```{r}
GCFID_Results_uncalibrated <- map(GCFID_Results_calibrated, function(df) {
  # Build surrogate calibration look up
  surrogate_params <- df %>%
    select(Compound, Sequence_ID) %>%
    distinct() %>%
    left_join(GCFID_quant, by = "Compound") %>%
    filter(!is.na(Response_surrogate)) %>%
    left_join(calibration_params, by = c("Response_surrogate" = "Compound", "Sequence_ID" = "Sequence_ID")) %>%
    select(Compound, Sequence_ID, Response_surrogate, Slope_surrogate = Slope, Intercept_surrogate = Intercept)

  # Apply surrogate calibration and retain slope/intercept
  df %>%
    left_join(surrogate_params, by = c("Compound", "Sequence_ID")) %>%
    mutate(
      Concentration_surrogate = (Area_ratio - Intercept_surrogate) / Slope_surrogate,
      Concentration_mgperL = if_else(
        is.na(Concentration_mgperL) | Concentration_mgperL == 0,
        Concentration_surrogate,
        Concentration_mgperL
      ),
      Concentration_mgperL = if_else(Concentration_mgperL < 0 | is.na(Concentration_mgperL), 0, Concentration_mgperL)
    )
})
```


## Excluding flagged compounds
```{r}
GCFID_Results_uncalibrated <- map(GCFID_Results_uncalibrated, function(df) {
  excluded_compounds <- GCFID_quant %>%
    filter(Exclude == "Yes") %>%
    pull(Compound)

  df %>%
    filter(!Compound %in% excluded_compounds)
})
```


## Joining results to Metadata, discard calibration information
Concentrations are in mgperL of hexane
```{r}
GCFID_Results_tidy <- map(GCFID_Results_uncalibrated, function(df) {
  df %>%
    select(
      Sequence_ID, Experiment_ID, `Sample Name`, Compound,
      Concentration_mgperL
    ) %>%
    pivot_wider(
      names_from = Compound,
      values_from = Concentration_mgperL,
      names_prefix = "Conc_"
    )
})

# Now join each tidy result to GCFID_meta
GCFID_Results_tidy <- map(GCFID_Results_tidy, function(df) {
  df %>%
    left_join(GCFID_meta, by = c(
      "Sequence_ID" = "Sequence_ID",
      "Experiment_ID" = "Experiment_ID",
      "Sample Name" = "Injection_ID"
    ))
})

# NA are treated like 0
GCFID_Results_tidy <- map(GCFID_Results_tidy, function(df) {
  df %>%
    mutate(across(starts_with("Conc_"), ~replace_na(.x, 0)))
})
```


## Calculating proportion of each compound and total FAME (g/dry g biomass)
```{r}
GCFID_Results_final <- map(GCFID_Results_tidy, function(df) {
  df %>%
    mutate(
      Total_FAME_mgperL = rowSums(select(., starts_with("Conc_")), na.rm = TRUE),
      Total_FAME_mg = Total_FAME_mgperL * ExtractantUsed_mL/1000,
      Total_FAME_g_per_g_biomass = Total_FAME_mg / (InitialSampleMass_mg)
    ) %>%
    mutate(across(starts_with("Conc_"), ~ .x / Total_FAME_mgperL, .names = "Prop_{.col}"))
})

GCFID_Results_all <- bind_rows(GCFID_Results_final)
```


# CLEAN-UP & EXPORT

## Selecting only important columns and giving more appealing compound names
STOP AT THIS POINT AND CHECK THE QCs if you want by running **View(GCFID_Results_all_neat)** in the console
```{r}

# Get all proportion columns
prop_cols <- names(GCFID_Results_all)[startsWith(names(GCFID_Results_all), "Prop_Conc_")]

# Extract the suffix from each column name
prop_suffixes <- sub("^Prop_Conc_", "", prop_cols)

# Create a data frame to match suffixes to clean labels
rename_df <- data.frame(
  original_col = prop_cols,
  FAME_ID = prop_suffixes,
  stringsAsFactors = FALSE
)

# Join with GCFID_labels to get clean names
rename_df <- merge(rename_df, GCFID_labels, by = "FAME_ID", all.x = TRUE)

# Create renaming vector
prop_rename_vector <- setNames(rename_df$FAME_label, rename_df$original_col)

# Select metadata, total FAME, and proportion columns
meta_cols <- c("Sequence_ID", "Experiment_ID", "Sample Name", "Sample_ID", 
               "Date_Collected", "KineticTime_d", "Reactor_ID", "Source", 
               "Block", "Date_analyzed", "Vial_tag", 
               "Total_FAME_g_per_g_biomass")

GCFID_Results_all_neat <- GCFID_Results_all[, c(meta_cols, prop_cols)]

# Rename proportion columns to clean FAME labels
names(GCFID_Results_all_neat)[names(GCFID_Results_all_neat) %in% names(prop_rename_vector)] <- 
  prop_rename_vector[names(GCFID_Results_all_neat)[names(GCFID_Results_all_neat) %in% names(prop_rename_vector)]]

# Replace NA with 0 in all proportion columns
prop_labels <- rename_df$FAME_label  # these are the new column names
GCFID_Results_all_neat[, prop_labels] <- lapply(
  GCFID_Results_all_neat[, prop_labels],
  function(x) ifelse(is.na(x), 0, x)
)

# Adjust dates format
GCFID_Results_all_neat <- GCFID_Results_all_neat %>%
  mutate(
    Date_Collected = as.Date(Date_Collected),
    Date_analyzed = as.Date(Date_analyzed)
  )
```


## Ordering, final check-up, clean-up, and export tidy dataset
Some alerts will be displayed here for  obvious issues in certain samples
[  ]USER INPUT (Type TRUE or FALSE for the variable DISCARD below). TRUE will discard flagged samples based on:
 1) FAME content higher than 100% (this is due to a problem when adding IS during the transesterification procedure)
 2) Methyl palmitate (C16:0) percentage of total FAME lower than 1 % (this signals a cultivation or sample prep outlier)
```{r}

# USER: DO YOU WANT TO DISCARD FLAGGED SAMPLES?
DISCARD <- TRUE




# Clean and reorder dataset
GCFID_Results_all_neat <- GCFID_Results_all_neat %>%
  rename(Injection_ID = `Sample Name`) %>%
  select(-Source, -Vial_tag) %>%
  select(Experiment_ID, Sequence_ID, Injection_ID, Sample_ID, Date_Collected, Date_analyzed, 
         Reactor_ID, KineticTime_d, Block, Total_FAME_g_per_g_biomass, everything()) %>%
  filter(Block != "QC")

# Define FAME columns (starting from column 10 onward)
fame_columns <- names(GCFID_Results_all_neat)[10:length(names(GCFID_Results_all_neat))]

# Identify samples with issues
fame_issue <- GCFID_Results_all_neat %>%
  filter(Total_FAME_g_per_g_biomass > 1)

cultivation_issue <- GCFID_Results_all_neat %>%
  filter(`Methyl palmitate (C16:0)` < 0.005)

# Combine all flagged samples by Sample_ID and Experiment_ID
flagged_samples <- bind_rows(fame_issue, cultivation_issue) %>%
  distinct(Sample_ID, Experiment_ID)

# Print warnings
if (nrow(fame_issue) > 0) {
  warning(paste0(
    "⚠️ Internal standard issue detected in ", nrow(fame_issue), " sample(s): ",
    paste(paste(fame_issue$Sample_ID, fame_issue$Experiment_ID, sep = " (Exp: "), collapse = "), "),
    ")",
    "\n"
  ))
}

if (nrow(cultivation_issue) > 0) {
  warning(paste0(
    "⚠️ Potential cultivation issue detected in ", nrow(cultivation_issue), " sample(s): ",
    paste(paste(cultivation_issue$Sample_ID, cultivation_issue$Experiment_ID, sep = " (Exp: "), collapse = "), "),
    ")",
    "\n"
  ))
}

# Create a flag column for samples to discard using both Sample_ID and Experiment_ID
GCFID_Results_all_neat <- GCFID_Results_all_neat %>%
  left_join(flagged_samples %>% mutate(flagged = TRUE),
            by = c("Sample_ID", "Experiment_ID")) %>%
  mutate(flagged = if_else(is.na(flagged), FALSE, flagged))

# Conditionally replace FAME values with NA
if (DISCARD) {
  GCFID_Results_all_neat <- GCFID_Results_all_neat %>%
    mutate(across(all_of(fame_columns), ~ if_else(flagged, NA_real_, .))) %>%
    select(-flagged)
  message("✅ FAME values in flagged samples were replaced with NA.")
} else {
  GCFID_Results_all_neat <- GCFID_Results_all_neat %>%
    select(-flagged)
  message("ℹ️ Flagged samples and their FAME values were retained.")
}

```

## Exporting dataset
```{r}
# Write the dataset to Excel
write_xlsx(GCFID_Results_all_neat, file.path(Export_dir, "GCFID_Results_Cleaned.xlsx"))
message(paste("✅ Cleaned dataset exported to:", file.path(Export_dir, "GCFID_Results_Cleaned.xlsx")))
```


# GENERAL VISUALIZATION

## Total FAME with non-parametric statistical test
```{r echo=TRUE, message=TRUE, warning=FALSE}

# Perform Kruskal-Wallis test for each Experiment_ID
kruskal_results <- GCFID_Results_all_neat %>%
  group_by(Experiment_ID) %>%
  summarise(
    kruskal = list(kruskal.test(Total_FAME_g_per_g_biomass ~ Block)),
    .groups = "drop"
  )

# Print results for each experiment
for (i in seq_len(nrow(kruskal_results))) {
  cat("\nExperiment:", kruskal_results$Experiment_ID[i], "\n")
  print(kruskal_results$kruskal[[i]])
}

# Ensure Block is a factor
GCFID_Results_all_neat$Block <- as.factor(GCFID_Results_all_neat$Block)

# Create faceted boxplot with jitter and p-value bar
p <- ggplot(GCFID_Results_all_neat, aes(x = Block, y = Total_FAME_g_per_g_biomass, fill = Block)) +
  geom_boxplot(outlier.shape = NA) +  # Hide outliers so jitter shows clearly
  geom_jitter(width = 0.2, alpha = 0.6, size = 1.5) +  # Add jittered points
  theme_minimal() +
  labs(title = "Total FAME by Block (Faceted by Experiment)",
       x = "Block",
       y = "Total FAME (g/g dry biomass)") +
  theme(legend.position = "none") +
  facet_wrap(~ Experiment_ID, scales = "free_x") +
  stat_compare_means(method = "kruskal.test", label = "p.format", 
                     label.y = max(GCFID_Results_all_neat$Total_FAME_g_per_g_biomass) * 1.05)

print(p)

# Export plot as JPEG to Export_dir
ggsave(filename = file.path(Export_dir, "Total_FAME_by_Block_Faceted.jpeg"),
       plot = p,
       width = 10, height = 6, dpi = 300)

message(paste("✅ Faceted boxplot with jitter exported to:", 
              file.path(Export_dir, "Total_FAME_by_Block_Faceted.jpeg")))

```

## Plotting the percentage of total FAME
```{r echo=TRUE}

# Identify FAME columns
fame_cols <- names(GCFID_Results_all_neat)[11:length(names(GCFID_Results_all_neat))]

# Reshape to long format
fame_long <- tidyr::pivot_longer(
  GCFID_Results_all_neat,
  cols = dplyr::all_of(fame_cols),
  names_to = "FAME_Type",
  values_to = "FAME_Percent"
)

# Compute global median and reorder factor levels explicitly (ascending)
fame_order <- fame_long %>%
  group_by(FAME_Type) %>%
  summarise(median_val = median(FAME_Percent, na.rm = TRUE)) %>%
  arrange(median_val) %>%   # ascending order
  pull(FAME_Type)

# Apply this order to the factor
fame_long$FAME_Type <- factor(fame_long$FAME_Type, levels = fame_order)

# Create the plot
p <- ggplot(fame_long, aes(x = FAME_Percent, y = FAME_Type, fill = Block)) +
  geom_boxplot(outlier.size = 0.8, alpha = 0.6) +
  facet_grid(rows = vars(Experiment_ID), scales = "free_y", space = "free_y") +
  scale_x_continuous(labels = scales::percent_format(scale = 100)) +  # Convert fractions to %
  theme_bw() +
  theme(
    strip.text.y = element_text(angle = 0),
    axis.text.y = element_text(size = 8),
    panel.spacing = grid::unit(0.5, "lines")
  ) +
  labs(
    title = "Horizontal Boxplot of FAME% by Experiment_ID (Ordered Ascending by Median Abundance)",
    x = "Percentage of total FAME",
    y = "FAME species"
  )

print(p)

# Export plot
ggsave(
  filename = file.path(Export_dir, "Profile_FAME_Boxplot_by_Experiment_Ordered.jpeg"),
  plot = p,
  width = 12,
  height = 22,
  dpi = 400
)

message(paste("✅ Ascending ordered FAME% boxplot (with percentage axis) exported to:",
              file.path(Export_dir, "FAME_Boxplot_by_Experiment_Ordered.jpeg")))

```
